---
# title: Mapping with **`ggplot2`**
format:
  revealjs:
    theme: FACE-IT_pres.scss
    self-contained: true
    transition: slide
    scrollable: true
editor: source
---

# Mapping with **`ggplot2`** {background-color="#008980" style="text-align: center"}

*Robert Schlegel*

```{r maps-1-opts, echo=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  echo = TRUE
)
```

## Problem

* Where do we start?
* How do we show scale?
* What if we need to inset a map?

## Solution

* **`tidyverse`** contains a wealth of mapping support
* **`ggsn`** provides code for scale bars etc.
* Maps are created the same as basic plots, so can be combined

## Setup

```{r map-load}
library(tidyverse) # Contains most of what we need

library(ggpubr) # For combining figures

library(ggsn) # Contains code to make scale bars

library(palmerpenguins) # Data used in an example
```

## Default map

```{r}
earth_1 <- ggplot() +
  borders(fill = "grey70", colour = "black") + # The global shape file
  coord_equal() # Equal sizing for lon/lat 
earth_1
```
## Cropping

```{r}
green_1 <- ggplot() +
  borders(fill = "grey70", colour = "black") +
  coord_equal(xlim = c(-75, -10), ylim = c(58, 85)) # Force lon/lat extent
green_1
```

## Regions

```{r}
map_data('world') %>% 
  select(region) %>% 
  distinct() %>% 
  arrange(region)
```

## Direct access

```{r}
map_data_green <- map_data('world') %>% 
  filter(region == "Greenland")
```

## A new concept?

-   A basic map is the same as a scatterplot with an x and y axis. 
-   Compare a dot plot of penguin weight data against a dot plot of the coastline of Greenland.

::: columns
::: {.column}
Penguin dots:

```{r}
ggplot(data = penguins, 
       aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point()
```
:::

::: {.column}
Greenland coast dots:

```{r}
ggplot(data = map_data_green, 
       aes(x = long, y = lat)) +
  geom_point()
```
:::
:::

## Polygons

-   Dots are not the ideal method for creating maps
-   For that we want to use polygons
-   Be careful with these, they can be a bit tricky

## Polygons

```{r}
green_2 <- ggplot(data = map_data_green, aes(x = long, y = lat)) +
  geom_polygon(aes(group = group), # What is this doing?
               fill = "green", colour = "black") # Note these are outside of aes() 
green_2
```

## Specific labels

-   A map is almost always going to need some labels and other visual cues. 
-   The following code chunk shows how this differs if we want to add just one label at a time. 
-   This can be useful if each label needs to be different from all other labels for whatever reason. 
-   We may also see that the text labels we are creating have `\n` in them. 
-   When R sees these two characters together like this it reads this as an instruction to return down a line.

## Specific labels

```{r}
green_3 <- green_2 +
    annotate("text", label = "Greenland", 
           x = -40, y = 75.0, 
           size = 7.0, 
           fontface = "bold.italic") +
  annotate("text", label = "North\nAtlantic\nOcean", 
           x = -20, y = 64.0, 
           size = 5.0, 
           angle = 330, 
           colour = "navy") +
  annotate("label", label = "Baffin\nBay", 
           x = -62, y = 70, 
           size = 5.0,
           fill = "springgreen") +
  annotate("segment", 
           x = -50, xend = -30, y = 73, yend = 73)
green_3
```

## Scale bars

-   There is no default scale bar function in the **`tidyverse`**
-   **`ggsn`** is devoted to adding scale bars and North arrows for **`ggplot2`** 
-   There are heaps of options so we'll just focus on one of them for now
-   We may see the list of available North arrow shapes by running `northSymbols()`.

```{r}
green_4 <- green_3 +
  scalebar(data = map_data_green, location = "bottomleft", # Set location of bar,
           dist = 500, dist_unit = "km", transform = TRUE, # Size of scale bar
           st.size = 4, height = 0.03, st.dist = 0.04) # Set particulars
green_4
```

## Insetting

-   To inset a map we must learn how to create a 'grob'
-   **`ggplot2`** objects are different from normal objects (i.e. dataframes)
-   They have their own way of storing and accessing data
-   To convert to the **`ggplot2`** format we convert it into a 'grob' 
-   Once converted, we may then plop it onto our figure/map wherever we please
-   Both of these steps are accomplished with the single function `annotation_custom()`
-   This is also a good way to add logos or any other sort of image to a map/figure
-   It's even possible to add GIFs (see Bonus content)

## Insetting

```{r}
earth_2 <- earth_1 + theme_void()
earth_2
```

## Insetting

```{r}
green_5 <- green_4 +
  annotation_custom(grob = ggplotGrob(earth_2), # Convert the earth plot to a grob
                    xmin = -30, xmax = -10,
                    ymin = 76, ymax = 84)
green_5
```

## Rounding it out

```{r}
green_final <- green_5 +
  scale_x_continuous(breaks = seq(-60, -20, 20),
                     labels = c("60°W", "40°W", "20°w"),
                     position = "bottom") +
  scale_y_continuous(breaks = seq(60, 80, 10),
                     labels = c("60°N", "70°N", "80°N"),
                     position = "right") +
  labs(x = "", y = "")
green_final
```

## Rounding it out

-   And lastly we save the fruits of our labours.

```{r maps-final-save, eval=FALSE}
ggsave(plot = sa_final, filename = "figures/green_final.pdf", 
       height = 6, width = 8)
```

