---
# title: Domesticating data
format:
  revealjs:
    theme: FACE-IT_pres.scss
    self-contained: true
    transition: slide
editor: source
---

# Domesticating data {background-color="#008980" style="text-align: center"}

*Robert Schlegel*

```{r tidy-3-opts, echo=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  echo = TRUE
)
```

## Problem

-   How do we summarise our data if they have multiple categories?
<<<<<<< HEAD
=======
-   What are the limitations of this approach?
-   Does this really help with our workflow?
>>>>>>> fbf7096585c5ec67d4d40972a54a62933968c7f0

## Solution

-   `group_by()` provides powerful options for data analysis
<<<<<<< HEAD

## Setup

```{r}
library(tidyverse) # All-in-one

sst_NOAA <- read_csv("course_material/data/sst_NOAA.csv") # SST data
=======
-   The only limits are our imagination, but be cautious
-   We end with the example exercise from Day 1

## Setup

```{r}
library(tidyverse) # All-in-one

library(lubridate) # Working with dates

sst_NOAA <- read_csv("../data/sst_NOAA.csv") # SST data
>>>>>>> fbf7096585c5ec67d4d40972a54a62933968c7f0
```

## The next level

In the previous session we covered the five main transformation functions one would use in a typical **tidy** workflow. But to really unlock their power we need to learn how to use them with `group_by()`. This is how we may calculate statistics based on the different grouping variables within our data, such as sites or species or months.

## `group_by()`

- Note that this function will not appear to do anything by itself
- This can cause issues if we aren't paying attention

```{r}
sst_NOAA_site <- sst_NOAA %>% group_by(site)
```

::: columns
::: {.column}
```{r}
sst_NOAA %>% head()
<<<<<<< HEAD
```
:::

::: {.column}
```{r}
sst_NOAA_site %>% head()
```
:::
:::

## `group_by()`

-   But when we `summarise()` the data...

::: columns
::: {.column}
```{r}
sst_NOAA %>% 
  summarise(mean_temp = mean(temp, na.rm = TRUE)) %>% 
  head()
```
:::

::: {.column}
```{r}
sst_NOAA_site %>% 
  summarise(mean_temp = mean(temp, na.rm = TRUE)) %>% 
  head()
```
:::
:::

## `ungroup() `

-   One must explicitly tell R to remove a group

```{r}
sst_NOAA_ungroup <- sst_NOAA_site %>% ungroup()
=======
```
:::

::: {.column}
```{r}
sst_NOAA_site %>% head()
>>>>>>> fbf7096585c5ec67d4d40972a54a62933968c7f0
```
:::
:::

<<<<<<< HEAD
::: columns
::: {.column}
```{r}
sst_NOAA_site %>% 
  summarise(mean_temp = mean(temp, na.rm = TRUE)) %>% 
  head()
```
:::

::: {.column}
```{r}
sst_NOAA_ungroup %>% 
  summarise(mean_temp = mean(temp, na.rm = TRUE)) %>% 
  head()
```
:::
:::

## Multiple groups

-   As one may have guessed by now, grouping is not confined to a single column
-   One may use any number of columns to perform elaborate grouping measures


```{r, eval=FALSE}
# Create groupings based on temperatures
sst_NOAA_temp_group <- sst_NOAA %>% 
  group_by(round(temp))

# Create groupings based on site and month
sst_NOAA_temp_month_group <- sst_NOAA %>% 
  mutate(month = month(t)) %>% 
  group_by(site, month)
=======
## `group_by()`

-   But when we `summarise()` the data...

::: columns
::: {.column}
```{r}
sst_NOAA %>% 
  summarise(mean_temp = mean(temp, na.rm = TRUE)) %>% 
  head()
```
:::

::: {.column}
```{r}
sst_NOAA_site %>% 
  summarise(mean_temp = mean(temp, na.rm = TRUE)) %>% 
  head()
>>>>>>> fbf7096585c5ec67d4d40972a54a62933968c7f0
```
:::
:::

<<<<<<< HEAD
## Chain functions

-   Generally we do not group objects separately
-   Grouping is performed within code chunks

```{r}
sst_NOAA_site_mean <- sst_NOAA %>% # Choose a base dataframe
  group_by(site) %>% # Group by the site column
  summarise(mean_temp = mean(temp, na.rm = TRUE), # Calculate means
            count = n() # Count observations
            ) # Safety measure
sst_NOAA_site_mean
```

## Grouped transformations

We've played around quite a bit with grouping and summarising, but that's not all we can do. We can use `group_by()` very nicely with `filter()` and `mutate()` as well. Not so much with `arrange()` and `select()` as these are designed to work on the entire dataframe at once, without any subsetting.

We can do some rather imaginative things when we combine all of these tools together. In fact, we should be able to accomplish almost any task we can think of. For example, what if we wanted to create a new object that was a subset of only the sites in the `SACTN` that had at least 30 years (360 months) of data?

```{r tidiest-group-6, eval=FALSE}
SACTN_30_years <- SACTN %>%
  group_by(site, src) %>%
  filter(n() > 360) %>% 
  ungroup()
```

Or what if we wanted to calculate anomaly data for each site?

```{r tidiest-group-7, eval=FALSE}
SACTN_anom <- SACTN %>%
  group_by(site, src) %>% 
  mutate(anom = temp - mean(temp, na.rm = T)) %>% 
  select(site:date, anom, depth, type) %>% 
  ungroup()
=======
## `ungroup() `

-   One must explicitly tell R to remove a group

```{r}
sst_NOAA_ungroup <- sst_NOAA_site %>% ungroup()
>>>>>>> fbf7096585c5ec67d4d40972a54a62933968c7f0
```

::: columns
::: {.column}
```{r}
sst_NOAA_site %>% 
  summarise(mean_temp = mean(temp, na.rm = TRUE)) %>% 
  head()
```
:::

::: {.column}
```{r}
sst_NOAA_ungroup %>% 
  summarise(mean_temp = mean(temp, na.rm = TRUE)) %>% 
  head()
```
:::
:::

## Multiple groups

-   As one may have guessed by now, grouping is not confined to a single column
-   One may use any number of columns to perform elaborate grouping measures


```{r, eval=FALSE}
# Create groupings based on temperatures
sst_NOAA_temp_group <- sst_NOAA %>% 
  group_by(round(temp))

# Create groupings based on site and month
sst_NOAA_temp_month_group <- sst_NOAA %>% 
  mutate(month = month(t)) %>% 
  group_by(site, month)
```

<<<<<<< HEAD
As one may imagine, performing intricate logical arguments like this may get out of hand rather quickly. It is advisable to save intermediate steps in a complex workflow to avoid too much heartache. Where exactly these 'fire breaks' should be made is up to the person writing the code.

<!-- ## Expand -->

<!-- ## Applied stats -->

## Pipe into **`ggplot2`**

It is also possible to combine piped code chunks and **`ggplot2`** chunks into one 'combi-chunk'. I prefer not to do this as I like saving the new dataframe I have created as an object in my environment before visualising it so that if anything has gone wrong (as things tend to do) I can more easily find the problem.
=======
## Chain functions
>>>>>>> fbf7096585c5ec67d4d40972a54a62933968c7f0

-   Generally we do not group objects separately
-   Grouping is performed within code chunks

```{r}
<<<<<<< HEAD
SACTN %>% # Choose starting dataframe
  filter(site %in% c("Bordjies", "Tsitsikamma", "Humewood", "Durban")) %>% # Select sites
  select(-depth, -type) %>% # Remove depth and type columns
  mutate(month = month(date), # Create month column
         index = paste(site, src, sep = "/ ")) %>% # Create individual site column
  group_by(index, month) %>% # Group by individual sites and months
  summarise(mean_temp = mean(temp, na.rm = TRUE), # Calculate mean temperature
            sd_temp = sd(temp, na.rm = TRUE)) %>% # Calculate standard deviation
  ggplot(aes(x = month, y = mean_temp)) + # Begin with ggplot, switch from '%>%' to '+'
  geom_ribbon(aes(ymin = mean_temp - sd_temp, ymax = mean_temp + sd_temp), 
              fill = "black", alpha = 0.4) + # Create a ribbon
  geom_line(col = "red", size = 0.3) + # Create lines within ribbon
  facet_wrap(~index) + # Facet by individual sites
  scale_x_continuous(breaks = seq(2, 12, 4)) + # Control x axis ticks
  labs(x = "Month", y = "Temperature (°C)") + # Change labels
  theme_dark() # Set theme
=======
sst_NOAA_site_mean <- sst_NOAA %>% # Choose a base dataframe
  group_by(site) %>% # Group by the site column
  summarise(mean_temp = mean(temp, na.rm = TRUE), # Calculate means
            count = n()) # Count observations
sst_NOAA_site_mean
>>>>>>> fbf7096585c5ec67d4d40972a54a62933968c7f0
```

## Grouped transformations

We've played around quite a bit with grouping and summarising, but that's not all we can do. We can use `group_by()` very nicely with `filter()` and `mutate()` as well. Not so much with `arrange()` and `select()` as these are designed to work on the entire dataframe at once, without any subsetting. We can do some rather imaginative things when we combine all of these tools together. In fact, we should be able to accomplish almost any task we can think of.

<<<<<<< HEAD
## Rename variables (columns) with `rename()`
=======
## Examples {auto-animate=true}
>>>>>>> fbf7096585c5ec67d4d40972a54a62933968c7f0

-   Filter sites that don't have a max temperature above 20°C

```{r}
sst_NOAA_20 <- sst_NOAA %>%
  group_by(site) %>%
  filter(max(temp) > 20) %>% 
  ungroup()
unique(sst_NOAA_20$site)
```

<<<<<<< HEAD
## Create a new dataframe for a newly created variable (column) with `transmute()`
=======
## Examples {auto-animate=true}
>>>>>>> fbf7096585c5ec67d4d40972a54a62933968c7f0

-   Calculate anomalies for each site

```{r}
sst_NOAA_anom <- sst_NOAA %>%
  group_by(site) %>% 
  mutate(anom = temp - mean(temp, na.rm = T)) %>%
  ungroup()
head(sst_NOAA_anom)
```
## Examples {auto-animate=true}

<<<<<<< HEAD
## Count observations (rows) with `n()`

We have already seen this function sneak it's way into a few of the code chunks in the previous session. We use `n()` to count any grouped variable automatically. It is not able to be given any arguments, so we must organise our dataframe in order to satisfy it's needs. It is the diva function of the **`tidyverse`**; however, it is terribly useful as we usually want to know how many observations our summary stats are based. First we will run some stats and create a figure without documenting `n`. Then we will include `n` and see how that changes our conclusions.
=======
-   Calculate mean and standard deviations for two sites
>>>>>>> fbf7096585c5ec67d4d40972a54a62933968c7f0

```{r}
sst_NOAA %>% 
  filter(site == "Med" | site == "WA") %>%
  group_by(site) %>% 
  summarise(mean_temp = mean(temp, na.rm = TRUE), 
            sd_temp = sd(temp, na.rm = TRUE))
```

## Examples {auto-animate=true}

-   Calculate mean and standard deviations for two sites

<<<<<<< HEAD
```{r  tidiest-n-3, fig.cap="Frequency distribution of mean temperature for each time series in the SACTN dataset."}
SACTN %>% 
  group_by(site, src) %>% 
  summarise(mean_temp = round(mean(temp, na.rm = T))
            ) %>% 
  ungroup() %>% 
  ggplot(aes(x = mean_temp)) +
  geom_density(fill = "seagreen", alpha = 0.6) +
  labs(x = "Temperature (°C)")
```

### Select observations (rows) by number with `slice()`
=======
```{r}
# First create a character vector containing the desired sites
selected_sites <- c("Med", "WA")

# Then calculate the statistics
sst_NOAA %>% 
  filter(site %in% selected_sites) %>%
  group_by(site) %>% 
  summarise(mean_temp = mean(temp, na.rm = TRUE), 
            sd_temp = sd(temp, na.rm = TRUE))
```

## Examples {auto-animate=true}

-   Only days with temperatures above 10°C and below 15°C
>>>>>>> fbf7096585c5ec67d4d40972a54a62933968c7f0

::: columns
::: {.column}
```{r}
sst_NOAA %>% 
  filter(site == "Med", 
         temp > 10, temp < 15) %>% 
  nrow()
```
:::

<<<<<<< HEAD
## Summary functions

There is a near endless sea of possibilities when one starts to become comfortable with writing R code. We have seen several summary functions used thus far. Mostly in straightforward ways. But that is one of the fun things about R, the only limits to what we may create are within our mind, not the program. Here is just one example of a creative way to answer a straightforward question: 'What is the proportion of recordings above 15°C per source?'. Note how we may refer to columns we have created within the same chunk. There is no need to save the intermediate dataframes if we choose not to.
    
```{r tidiest-summary}
SACTN %>% 
  na.omit() %>% 
  group_by(src) %>%
  summarise(count = n(), 
            count_15 = sum(temp > 15)) %>% 
  mutate(prop_15 = count_15/count) %>% 
  arrange(prop_15)
=======
::: {.column}
```{r}
sst_NOAA %>% 
  filter(site == "Med", 
         !(temp <= 10 | temp  >= 15)) %>% 
  nrow()
>>>>>>> fbf7096585c5ec67d4d40972a54a62933968c7f0
```
:::
:::

## The new age *redux*

```{r tidiest-new-age}
read_csv("../data/sst_NOAA.csv") %>% # Load the SACTN Day 1 data
  mutate(month = month(t, label = T)) %>% # Then create a month abbreviation column
  group_by(site, month) %>% # Then group by sites and months
  summarise(mean_temp = mean(temp, na.rm = TRUE), # Lastly calculate the mean
            sd_temp = sd(temp, na.rm = TRUE)) %>% # and the SD
  ggplot(aes(x = month, y = mean_temp, group = site)) + # Begin ggplot
  geom_ribbon(aes(ymin = mean_temp - sd_temp, ymax = mean_temp + sd_temp), 
              fill = "black", alpha = 0.4) + # Create a ribbon
  geom_point(aes(colour = site)) + # Create dots
  geom_line(aes(colour = site, group = site)) + # Create lines
  labs(x = "Month", y = "Temperature (°C)", colour = "Site") # Change labels
<<<<<<< HEAD
```
=======
```

## Summary functions

There is a near endless sea of possibilities when one starts to become comfortable with writing R code. We have seen several summary functions used thus far. Mostly in straightforward ways. But that is one of the fun things about R, the only limits to what we may create are within our mind, not the program. Here is just one example of a creative way to answer a straightforward question: 'What is the proportion of recordings above 20°C per source?'. Note how we may refer to columns we have created within the same chunk. There is no need to save the intermediate dataframes if we choose not to.
    
## Summary functions

```{r tidiest-summary}
sst_NOAA %>%  
  group_by(site) %>%
  summarise(count = n(), 
            count_15 = sum(temp > 20)) %>% 
  mutate(prop_15 = count_15/count) %>% 
  arrange(prop_15)
```
>>>>>>> fbf7096585c5ec67d4d40972a54a62933968c7f0
