---
# title: Local data
format:
  revealjs:
    theme: FACE-IT_pres.scss
    self-contained: true
    transition: slide
editor: source
---

# Local data {background-color="#008980" style="text-align: center"}

*Robert Schlegel*

```{r tidy-3-opts, echo=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  echo = TRUE
)
```

## Problem

-   Now that we can access data, how do we manage them?
-   Are there tools to help us with this?
-   How can we perform more complex analyses?

## Solution

-   We will go more in depth into advanced workflows
-   The most commonly used tools are discussed
-   A few examples of applied statistics are provided

## Setup

```{r}
library(tidyverse) # All-in-one
```

## rasters

<!-- Manually download data from Bio-oracle and load into R -->

## Shape files

<!-- Show how to load them and plot them -->

## NetCDF

## Expand

<!-- Show how to fill in the gaps in a dataset -->
<!-- DO this by using geom_line in a time series -->

## tidy statistics

<!-- Perform a posthoc ANOVA and show how to filter out non-significant results -->

## Grouped statistics

<!-- Show how to perform a grouped linear model -->

