---
# title: Mapping the Arctic
format:
  revealjs:
    theme: FACE-IT_pres.scss
    self-contained: true
    transition: slide
editor: source
---

# Mapping the Arctic {background-color="#008980" style="text-align: center"}

*Robert Schlegel*

```{r maps-3-opts, echo=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  echo = TRUE,
  eval = FALSE
)
```

## Problem

-   How do we plot maps near the poles more accurately?
-   What effect does this have on other map layers?
-   This is time intensive, is there a faster way?

## Solution

-   We must learn more about map projections
-   We will look at packages that help us project raster layers etc.
-   **`ggoceanmaps`** provides the full suite of solutions

## Setup

Now that we've seen how to produce complex maps with **`ggplot2`** we are going to learn how to bring Google maps into the mix. Some kind hearted person has made a package for R that allows us to do this relatively easily. But that means we will need to install another new package.

After we learn how to use Google maps we will take some time to learn additional mapping techniques that we may want for creating publication quality figures.

The package we will need for this tut is **`ggmap`**. It is a bit beefy so let's get started on installing it now. Once it's done we will activated the packages we need and load some site points for mapping.

```{r ggmap-load}
# Load libraries
library(tidyverse)

# Load data

```

Take a moment to look at the data we loaded. What does it show?

## Polar projections

## Projecting layers

## **`ggoceanmaps`**


