---
# title: Basics of **`ggplot2`**
format:
  revealjs:
    # logo: ../logo/FACE-IT_Logo_900.png
    # slide-number: true
    # show-slide-number: print
    # preview-links: true
    theme: FACE-IT.scss
    self-contained: true
    transition: slide
editor: source
---

# Basics of **`ggplot2`** {background-color="#008980" style="text-align: center"}

<!-- *Robert Schlegel* -->

## Problem

* Default R plotting (_Base R_) hasn't advanced much since the 90s
* Non-intuitive syntax, functions, and arguments
* Not enough control over final plot
* Published figures do not look very professional

## Solution

* The **`ggplot2`** package uses the [grammar of graphics](https://link.springer.com/book/10.1007/0-387-28695-0)
* Is integrated into the **`tidyverse`**
* Easier syntax with intuitive functions and arguments
* Massive range of well developed support and extensions

## Setup

We will need the following two packages for the examples in these slides.

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Load tidyverse - contains ggplot2
library(tidyverse)

# Load Palmer penguins - contains dataset
library(palmerpenguins)
```

## Basics

* One figure is made of one chunk of code
* Starts with `ggplot()`
* Each line is connected with a `+`
* Add shapes with `geom` functions
    * e.g. `geom_point()` adds points
* Plot skeleton created within `aes()`
* Arguments assigned like normal functions (e.g. `ggplot(data = penguins)`)

## Basic plot

```{r, echo=TRUE}
ggplot(data = penguins,
       aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(aes(colour = species))
```

## Focus on `aes()`

* Understand when to use `aes()`
* Columns from the data go inside `aes()`
    * e.g. `geom_point(aes(colour = species))`
* Static values go outside `aes()`
* Mistakes with `aes()` are common when learning **`ggplot2`** 
* Good starting point when looking for errors

## Inside `aes()` {auto-animate=true}

```{r, echo=TRUE, warning=FALSE, message=FALSE}
ggplot(data = penguins,
       aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(aes(colour = island)) # 'island' is a column from 'penguins'
```

## Outside `aes()` {auto-animate=true}

```{r, echo=TRUE, warning=FALSE, message=FALSE}
ggplot(data = penguins,
       aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(colour = "red") # Make all points red
```

## Inside outside? {auto-animate=true}

```{r, echo=TRUE, warning=FALSE, message=FALSE}
ggplot(data = penguins,
       aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(aes(colour = "red")) # What is happening here
```

## Inside outside? {auto-animate=true}

```{r, echo=TRUE, error=TRUE}
ggplot(data = penguins,
       aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(colour = species) # Why does this cause an error?
```

## More than just colour {auto-animate=true}

```{r, echo=TRUE, warning=FALSE, message=FALSE}
ggplot(data = penguins,
       aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(aes(size = flipper_length_mm, shape = island)) # What else can we add?
```

## Add a geom {auto-animate=true}

```{r, echo=TRUE, warning=FALSE, message=FALSE}
ggplot(data = penguins,
       # NB: Arguments passed to first aes() apply to all geoms
       aes(x = body_mass_g, y = bill_length_mm, colour = species)) +
  geom_point() +
  geom_smooth(method = "lm")
```
